---
title: "Bar plot of coverage results across MAGs for each sample"
author: "Coverage_Quant"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    keep_md: yes
---





```{r setup, include=FALSE}
# The following setting is important, do not omit.
knitr::opts_chunk$set(echo = TRUE,
                      fig.height = 8, 
                      fig.width = 8, 
                      dev = c("png", "pdf"),
                      dpi = 1000)
options(scipen = 999) #Prevent scientific notation
rm(list = ls()) # Clear environment
```

*Note: Open scripts.Rproj first, then script. To easily use relative paths, click the down button next to knit and then click "Knit Directory --> Project Directory". This should make loading and saving files much easier.*



# Coverage Bar Plots

Produce stacked bar plots of the MAG coverage across each sample.

<span style="color:red">**Red text**</span> indicates regions that require the user to modify.
Regardless, the user should check over all code blocks to ensure that everything is running correctly.



## 1. Load packages

Load packages.
```{r load_packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(maditr)
library(reshape2)
library(htmlwidgets)
library(plotly)
library(grid)
library(cowplot)
```



## 2. Set variables

<span style="color:red">**Change file names and condition variables where appropriate.**</span>
```{r variables}
# Input metadata file
metadata.file <- "mag_metadata.tsv"
# Input coverage file
cov.file        <- "coverage.tsv.gz"
read_stats.file <- "coverage.read_stats.tsv"
# Output coverage results
prefix <- "mag2scaffold"
out_ReadCount.file <- paste(prefix, "-ReadCount.tsv", sep='')
out_TPM.file       <- paste(prefix, "-TPM.tsv", sep='')
```



## 3. Load, clean, and pre-processing datasets

Load metadata file.
``` {r load_metadata}
metadata <- read.table(metadata.file, sep='\t', header=TRUE, check.names=FALSE, stringsAsFactors=FALSE, comment.char='')
```


Load (mapped) read stats file.
``` {r load_readstats}
read_stats <- read.table(read_stats.file, sep='\t', 
                         header=TRUE, check.names=FALSE, stringsAsFactors=FALSE) %>% 
  column_to_rownames("sample_id")
```


Load `coverm genome` MAG coverage results - clean-up sample names.
```{r load_cov}
cov <- read.table(cov.file, sep='\t',
                  header=TRUE, check.names=FALSE, stringsAsFactors=FALSE)
```


Extract coverage values from `coverm genome` results file and format into a matrix for plotting.
``` {r extract_coverage_values}
# Build matrix using: Group_ID<->sample
cov.ReadCount <- cov %>%
  dcast(Genome ~ Sample, value.var = "Read Count") %>%
  filter(!Genome %in% c("unmapped")) %>%
  column_to_rownames("Genome")

cov.TPM <- cov %>%
  dcast(Genome ~ Sample, value.var = "TPM") %>%
  filter(!Genome %in% c("unmapped")) %>%
  column_to_rownames("Genome")

cov.RPKM <- cov %>%
  dcast(Genome ~ Sample, value.var = "RPKM") %>%
  filter(!Genome %in% c("unmapped")) %>%
  column_to_rownames("Genome")
```


Add unaligned/unassigned read counts to matrix.
```{r add_unalinged_reads}
col.order <- colnames(cov.ReadCount)
cov.ReadCount["Unassigned",] = read_stats[col.order,]$total_reads - read_stats[col.order,]$mapped_reads

# Add entry to metadata file.
metadata <- rbind(metadata, c("Unassigned", "Unassigned"))
```


Colors for plotting.
```{r colors_for_plotting}
t <- metadata[,c("Label", "Label_color")] %>% 
  group_by(Label, Label_color) %>% 
  summarize(n=n(), .groups="drop")

colors <- t$Label_color
names(colors) <- t$Label

colors <- colors[ - which(names(colors) == "d__Viruses--r__")]
colors[["d__Viruses--r__"]] <- "#969696"

colors <- colors[ - which(names(colors) == "Unassigned")]
colors[["Unassigned"]] <- "#4d4d4d"
```


Merge `cov` matrices with metadata.
Control the order of the bars in the plot using the `names.ordered` list.
<span style="color:red">**Adjust `names.ordered` as required.**</span>
```{r merge_metadata}
names.ordered <- c(
  "Cmer_10D",
  "Cmer_10D_chloroplast",
  "Cmer_10D_mitochondria",
  "Gsulp_YNP5587_1_nuclear",
  "Gsulp_YNP5587_1_chloroplast",
  "Gsulp_YNP5587_1_mitochondria",
  "d__Archaea--p__Micrarchaeota",
  "d__Archaea--p__Nanoarchaeota",
  "d__Archaea--p__Thermoplasmatota",
  "d__Archaea--p__Thermoproteota",
  "d__Bacteria--p__Actinobacteriota",
  "d__Bacteria--p__Aquificota",
  "d__Bacteria--p__Bacteroidota",
  "d__Bacteria--p__Campylobacterota",
  "d__Bacteria--p__Chlamydiota",
  "d__Bacteria--p__Deinococcota",
  "d__Bacteria--p__Dependentiae",
  "d__Bacteria--p__Desulfobacterota_B",
  "d__Bacteria--p__Dormibacterota",
  "d__Bacteria--p__Firmicutes",
  "d__Bacteria--p__Firmicutes_B",
  "d__Bacteria--p__Firmicutes_E",
  "d__Bacteria--p__Nitrospirota_A",
  "d__Bacteria--p__Patescibacteria",
  "d__Bacteria--p__Planctomycetota",
  "d__Bacteria--p__Proteobacteria",
  "d__Bacteria--p__SZUA-79",
  "d__Bacteria--p__Thermotogota",
  "d__Eukaryota--d__Eukaryota",
  "d__Viruses--r__Adnaviria",
  "d__Viruses--r__Duplodnaviria",
  "d__Viruses--r__Inoviridae",
  "d__Viruses--r__Monodnaviria",
  "d__Viruses--r__Riboviria",
  "d__Viruses--r__Varidnaviria",
  "d__Viruses--r__environmental samples",
  "d__Viruses--r__unclassified",
  "d__Viruses--r__unclassified archaeal viruses",
  "d__Viruses--r__unclassified bacterial viruses",
  "d__Viruses--r__unclassified virophages",
  "d__Viruses--r__unclassified viruses",
  "d__Viruses--r__",
  "Unassigned"
)

cov.ReadCount <- merge(cov.ReadCount %>% as.matrix() %>% melt(), 
      metadata, 
      by.x = "Var1", by.y = "MAG_ID", 
      all.x = TRUE, all.y = FALSE) %>% 
  mutate(Label = factor(Label, levels = names.ordered)) %>%
  rename(MAG_ID = Var1) %>%
  rename(Sample = Var2)

cov.RPKM <- merge(cov.RPKM %>% as.matrix() %>% melt(), 
      metadata, 
      by.x = "Var1", by.y = "MAG_ID", 
      all.x = TRUE, all.y = FALSE) %>% 
  mutate(Label = factor(Label, levels = names.ordered)) %>%
  rename(MAG_ID = Var1) %>%
  rename(Sample = Var2)

cov.TPM <- merge(cov.TPM %>% as.matrix() %>% melt(), 
      metadata, 
      by.x = "Var1", by.y = "MAG_ID", 
      all.x = TRUE, all.y = FALSE) %>% 
  mutate(Label = factor(Label, levels = names.ordered)) %>%
  rename(MAG_ID = Var1) %>%
  rename(Sample = Var2)
```


Write read count tables to file (in case we want them downstream since they are now nicely formatted).
```{r write_matrices}
write.table(cov.ReadCount, out_ReadCount.file, 
            quote=FALSE, sep='\t', row.names=FALSE, col.names=TRUE)
write.table(cov.TPM, out_TPM.file, 
            quote=FALSE, sep='\t', row.names=FALSE, col.names=TRUE)
```



## 4. Plot ReadCounts

```{r barchart_ReadCounts}
p <- cov.ReadCount %>%
  ggplot(aes(x=Sample, y=value, fill=Label)) + 
    geom_bar(position="fill", stat="identity", aes(label=MAG_ID, label2=value)) +
    scale_fill_manual(values = colors) + 
    theme_classic() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# Write plot as interactive HTML file
saveWidget(suppressWarnings(ggplotly(p)), file=paste(prefix,".ReadCount.html", sep=''))

# Plot
p + theme(legend.position = "none")

# Plot legend
grid.newpage()
grid.draw(get_legend(p))
```


Plot Read Count with top most abundant samples highlighted.
```{r barchart_ReadCount_highlightedTop}
alpha.min <- 0.5
alpha.max <- 1.0
percent.cutoff <- 2

sample.sum <- cov.ReadCount %>% 
  group_by(Sample) %>% 
  summarize(sum=sum(value), .groups="drop") %>% 
  column_to_rownames("Sample")

d <- cov.ReadCount %>%
  mutate(proportion = ((value / sample.sum[Sample, ])*100)) %>%
  mutate(selected = if_else( proportion > percent.cutoff, TRUE, FALSE)) %>%
  mutate(selected_alpha = if_else(selected, 
                                  alpha.max, 
                                  alpha.min
                                  )) %>%
  mutate(selected_label = if_else(selected, 
                                  paste(MAG_ID, " (", round(proportion, 2), "%)", sep=''), 
                                  ""
                                  ))

# Write transformed data to file - Pivot to wide format for writing
t <- d %>%
  select(-selected_alpha, -selected_label) %>%
  pivot_wider(names_from = Sample,
              values_from = c(value, proportion, selected),
              names_glue = "{Sample} {.value}",
              names_vary="slowest",
              names_sort=TRUE) %>%
  mutate(all_selected = if_any(contains("selected"), ~ .x)) # Add column summarizing the selected columns (i.e., was the row selected at any "Sample")
write.table(t, file=paste(prefix,".ReadCount_highlightedTop.tsv", sep=''), sep='\t', quote=FALSE, row.names=FALSE, col.names=TRUE)

p <- d %>%
  ggplot(aes(x=Sample, y=value, 
             fill=Label, alpha=selected_alpha,
             color=selected, linetype=selected,
             label=MAG_ID, label2=value)) + 
  geom_bar(position="fill", stat="identity", linewidth=0.1) +
  geom_text(aes(label = selected_label), 
                position = position_fill(vjust = 0.5),
                size = 1
            ) +
  scale_fill_manual(values = colors) + 
  scale_alpha(range = c(alpha.min, alpha.max)) +
  scale_color_manual(values = c("black","black")) +
  scale_linetype_manual(values = c("blank", "solid")) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# Write plot as interactive HTML file
saveWidget(suppressWarnings(ggplotly(p)), file=paste(prefix,".ReadCount_highlightedTop.html", sep=''))

# Plot
p + theme(legend.position = "none")

# Plot legend
grid.newpage()
grid.draw(get_legend(p))
```



## 5. Plot TPM

```{r barchart_TPM}
p <- cov.TPM %>% 
  arrange(Label) %>%
  ggplot(aes(x=Sample, y=value, fill=Label)) + 
    geom_bar(position="stack", stat="identity", aes(label=MAG_ID, label2=value)) +
    scale_fill_manual(values = colors) + 
    theme_classic() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# Write plot as interactive HTML file
saveWidget(suppressWarnings(ggplotly(p)), file=paste(prefix,".TPM.html", sep=''))

# Plot
p + theme(legend.position = "none")

# Plot legend
grid.newpage()
grid.draw(get_legend(p))
```


Plot TPM with top most abundant samples highlighted.
```{r barchart_TPM_highlightedTop}
alpha.min <- 0.5
alpha.max <- 1.0
percent.cutoff <- 2

d <- cov.TPM %>%
  mutate(proportion = ((value/1000000)*100)) %>%
  mutate(selected = if_else( proportion > percent.cutoff, TRUE, FALSE)) %>%
  mutate(selected_alpha = if_else(selected, 
                                  alpha.max, 
                                  alpha.min
                                  )) %>%
  mutate(selected_label = if_else(selected, 
                                  paste(MAG_ID, " (", round(proportion, 2), "%)", sep=''), 
                                  ""
                                  ))

# Write transformed data to file - Pivot to wide format for writing
t <- d %>%
  select(-selected_alpha, -selected_label) %>%
  pivot_wider(names_from = Sample,
              values_from = c(value, proportion, selected),
              names_glue = "{Sample} {.value}",
              names_vary="slowest",
              names_sort=TRUE) %>%
  mutate(all_selected = if_any(contains("selected"), ~ .x)) # Add column summarizing the selected columns (i.e., was the row selected at any "Sample")
write.table(t, file=paste(prefix,".TPM_highlightedTop.tsv", sep=''), sep='\t', quote=FALSE, row.names=FALSE, col.names=TRUE)

p <- d %>%
  ggplot(aes(x=Sample, y=value, 
             fill=Label, alpha=selected_alpha,
             color=selected, linetype=selected,
             label=MAG_ID, label2=value)) + 
  geom_bar(position="fill", stat="identity", linewidth=0.1) +
  geom_text(aes(label = selected_label), 
                position = position_fill(vjust = 0.5),
                size = 1
            ) +
  scale_fill_manual(values = colors) + 
  scale_alpha(range = c(alpha.min, alpha.max)) +
  scale_color_manual(values = c("black","black")) +
  scale_linetype_manual(values = c("blank", "solid")) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# Write plot as interactive HTML file
saveWidget(suppressWarnings(ggplotly(p)), file=paste(prefix,".TPM_highlightedTop.html", sep=''))

# Plot
p + theme(legend.position = "none")

# Plot legend
grid.newpage()
grid.draw(get_legend(p))
```



## 6. Plot RPKM

```{r barchart_RPKM}
p <- cov.RPKM %>%
  ggplot(aes(x=Sample, y=value, fill=Label)) + 
    geom_bar(position="fill", stat="identity", aes(label=MAG_ID, label2=value)) +
    scale_fill_manual(values = colors) + 
    theme_classic() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# Write plot as interactive HTML file
saveWidget(suppressWarnings(ggplotly(p)), file=paste(prefix,".RPKM.html", sep=''))

# Plot
p + theme(legend.position = "none")

# Plot legend
grid.newpage()
grid.draw(get_legend(p))
```



Plot RPKM with top most abundant samples highlighted.
```{r barchart_RPKM_highlightedTop}
alpha.min <- 0.5
alpha.max <- 1.0
percent.cutoff <- 2

d <- cov.RPKM %>%
  mutate(proportion = ((value/1000000)*100)) %>%
  mutate(selected = if_else( proportion > percent.cutoff, TRUE, FALSE)) %>%
  mutate(selected_alpha = if_else(selected, 
                                  alpha.max, 
                                  alpha.min
                                  )) %>%
  mutate(selected_label = if_else(selected, 
                                  paste(MAG_ID, " (", round(proportion, 2), "%)", sep=''), 
                                  ""
                                  ))

# Write transformed data to file - Pivot to wide format for writing
t <- d %>%
  select(-selected_alpha, -selected_label) %>%
  pivot_wider(names_from = Sample,
              values_from = c(value, proportion, selected),
              names_glue = "{Sample} {.value}",
              names_vary="slowest",
              names_sort=TRUE) %>%
  mutate(all_selected = if_any(contains("selected"), ~ .x)) # Add column summarizing the selected columns (i.e., was the row selected at any "Sample")
write.table(t, file=paste(prefix,".RPKM_highlightedTop.tsv", sep=''), sep='\t', quote=FALSE, row.names=FALSE, col.names=TRUE)

p <- d %>%
  ggplot(aes(x=Sample, y=value, 
             fill=Label, alpha=selected_alpha,
             color=selected, linetype=selected,
             label=MAG_ID, label2=value)) + 
  geom_bar(position="fill", stat="identity", linewidth=0.1) +
  geom_text(aes(label = selected_label), 
                position = position_fill(vjust = 0.5),
                size = 1
            ) +
  scale_fill_manual(values = colors) + 
  scale_alpha(range = c(alpha.min, alpha.max)) +
  scale_color_manual(values = c("black","black")) +
  scale_linetype_manual(values = c("blank", "solid")) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# Write plot as interactive HTML file
saveWidget(suppressWarnings(ggplotly(p)), file=paste(prefix,".RPKM_highlightedTop.html", sep=''))

# Plot
p + theme(legend.position = "none")

# Plot legend
grid.newpage()
grid.draw(get_legend(p))
```





## 7. Session Info

```{r ressionInfo}
sessionInfo()
```